### 2020-05-27 writeup/intro.tex

%Multi-task learning is an inductive learning mechanism to improve generalization performance using related task data.
%Many state-of-the-art results in computer vision and natural language processing are obtained using multi-task learning.
Multi-task learning has recently become a powerful paradigm to solve complex prediction tasks in computer vision \cite{chexnet17,ZSSGM18}, natural language processing \cite{GLUE,superglue} and numerous other areas \cite{ZY17}.
%In multi-task learning, having related task data is fundamental to its performance.
%Multi-task learning is particularly powerful when there is limited labeled data for a task to be solved, meanwhile more labeled data from different but related tasks is available.
By combining multiple information sources, multi-task learning allows new information to be shared across different sources in a model \cite{C97}.
%For example, many applications in , and many other areas have been achieved by learning from multiple tasks together.
At the same time, it is well-known that the performance of multi-task learning methods depends on the relationship between the information sources.
\textit{Negative transfer}, the phenomenon where for a particular prediction task, multi-task learning performs worse than single-task learning, is prevalent when the information sources are heterogeneous \cite{AP16,BS17}.
While numerous studies have sought to alleviate negative transfer when it occurs \cite{YKGLHF20}, a rigorous understanding to the contributing causes of negative transfer has remained elusive in the literature \cite{R17}.
%This phenomenon, known as \textit{negative transfer}, is fundamental to the understanding of multi-task learning.
In this work, we consider learning multiple high-dimensional linear regression tasks to better understand when and why negative transfer happens. % for learning from multiple linear regression tasks.
We provide theoretical and practical insights to show how task data affects the transfer of information.
%We consider a setting where the target task has limited labeled data and show
%On the other hand, unless the structures across task data are well-understood, applying multi-task learning on several different datasets often result in suboptimal models (or negative transfer in more technical terms).

Identifying when negative transfer occurs requires comparing the performance of multi-task learning directly to single-task learning.
The technical challenge is to develop generalization bounds that are able to scale tightly with the qualities of task data, such as the number of datapoints.
In classical Rademacher or VC based theory of multi-task learning \cite{B00,AZ05,M06}, the generalization bounds are usually presented in a way so that the error goes down as more labeled data is added.
On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
More recent work has shown the benefit of learning multi-task representations for certain half-spaces \cite{MPR16} and multiple sparse regressions \cite{LPTV09,LPVT11}.

@@

In this work, we consider multiple high-dimensional linear regression tasks as input and focus on predicting a particular task that only has limited amount of labeled data.
Following \cite{HMRT19,BLLT20}, we assume that for every task $1\le i\le t$, its features are random vectors $x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
The label of $x$ is given by $y_i = x^{\top}\beta_i + \varepsilon$, where $\beta_i\in\real^p$ denotes the ground truth parameters for task $i$ and $\varepsilon$ denotes random noise with mean zero and variance $\sigma^2$.
Importantly, we assume that for the target task $t$, its training data size $n_t$ is a small constant $\rho_t$ times $p$. %\todo{rational}
%We shall assume that each task data follows a linear model, i.e. $y_i = X_i \beta_i + \varepsilon_i$, $1\le i\le k$.
%Here $\beta_i\in\real^p$ is the model parameter for the $i$-th task.
%Each row of $X_i\in\real^{n_i\times p}$ is assumed to be drawn i.i.d. from a fixed
%distribution with covariance matrix $\Sigma_i$.
We use a hard parameter sharing architecture that contains a shared body $B\in\real^{p\times r}$ for all tasks and a separate prediction head $\set{W_i \in \real^{r}}_{i=1}^t$ for each task \cite{R17,MTDNN19}.
%    \paragraph{Different covariates:}
This corresponds to minimizing the following objective.
\begin{align}
	\label{eq_mtl}
	f(B; W_1, \dots, W_t) = \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2.
\end{align}
%Note that we consider the natural parameterization without reweighting the tasks above.
%The shared body $B$ plays an important role because it allows information transfer between different task data.
%This is known as the hard parameter sharing architecture in the literature, where we control the capacity $r$ of $B$, e.g. \cite{KD12,WZR20}.
%We focus on comparing the test performance on a particular task solved with equation \eqref{eq_mtl} to single-task learning.
%The details are described in Algorithm \ref{alg_estimator}.
Let $\hat{\beta}_t^{\MTL} = B W_t$ denote the optimal predictor obtained from solving equation \eqref{eq_mtl} for task $t$.
Our goal is to compare the test error of $\hat{\beta}_t^{\MTL}$, denoted as $\te(\hat{\beta}_t^{\MTL})$, to $\te(\hat{\beta}_t^{\STL})$, where $\hat{\beta}_t^{\STL}$ is likewise obtained from equation \eqref{eq_mtl} but with task $t$ in isolation.

%\begin{figure}[!t]
%	\centering
%	\includegraphics[width=0.8\textwidth]{figures/model_distance_motivation.pdf}
%	\caption{Positive to negative transfer as the source task rotates further away from the target task (left to right).}
%	\label{fig_motivation}
%\end{figure}

\textbf{Main results.}
	We begin by observing that $\te(\hat{\beta}_t^{\MTL})$ can be decomposed into two parts, a variance part that is reduced from $\te(\hat{\beta}_t^{\STL})$, and a bias part that captures the difference between $\beta_t$ and the rest of $\beta$'s.
	We term the bias part as \textit{model shift bias}.
	Intuitively, whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ depends on the trade-off between the amount of variance reduced and the model shift bias part.
	For the high-dimensional regime when $p$ goes to infinity, we derive the asymptotic limit of $\te(\hat{\beta}_t^{\MTL}) - \te(\hat{\beta}_t^{\STL})$ as a function of the number of datapoints $\set{n_1, n_2}$, the covariance matrices $\set{\Sigma_1, \Sigma_2}$, the ground truth parameters $\set{\beta_1, \beta_2}$, and a certain fixed value derived from solving equation \eqref{eq_mtl} (see Theorem \ref{thm_main_informal} for the statement).
	Then, we show a similar result for any number of tasks that have the same covariates, i.e. the $X_i$'s are equal to each other.
	This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.

	While our focus is multi-task learning, our setup and technical tools are also applicable to transfer learning settings.
	In Section \ref{sec_taskonomy}, we provide a case study of the transfer function used in Taskonomy \cite{ZSSGM18}.
	We observe a similar trade-off between model-shift bias and variance for this setting.

	Based on the technical result, we provide theoretical insights over when multi-task learning performs better than single-task learning.
	\begin{itemize}
		\item First, we explain a sharp transition from positive to negative transfer as a parameter of \textit{task model dissimilarity} $\norm{\beta_1 - \beta_2}^2$.
	In Proposition \ref{prop_dist_transition}, we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.

		First, we provide a metric to determine positive versus negative transfer by comparing the test accuracies of single-task models.
		\item Second, we show that performing multi-task learning reduces the need for labeled data, which is a key finding of Taskonomy \cite{ZSSGM18}.
	We define the \textit{data efficiency ratio} as the smallest $\alpha\in(0, 1)$ such that if we only use an $\alpha$ fraction of labeled data, then the test error of $\hat{\beta}_t^{\MTL}$ matches that of the $\hat{\beta}_t^{\STL}$ on the entire set.
	In Proposition \ref{prop_data_efficiency}, we show that the data efficiency ratio of an illustrative example is at most $\alert{\frac {1}{2\rho_t}}$.

			Second, we validate that performing multi-task learning can reduce the need for labeled data.
	On 6 sentiment analysis tasks, we observe that just by using \alert{40\%} of the labeled data, the overall accuracy of multi-task learning matches that of learning every task in isolation.
		\item Finally, as the data size becomes more imbalanced between the source and target task, having mis-aligned task covariance matrices can result in suboptimal performance.
		In Proposition \ref{prop_covariate}, we show that as $n_1 / n_2$ becomes large, the best performing source task has have the same covariance matrix as the target task.
		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

		Finally, we show that as the data size between the source and target task becomes more imbalanced, aligning the covariances of the tasks becomes more beneficial.
	\end{itemize}

	A crucial technical tool that we develop for showing Theorem \ref{thm_main_informal} is the asymptotic limit of the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$ (for the setting of two tasks), which may be of independent interest.
	The result extends a well-known result on the trace of $(X_1^{\top}X_1)^{-1}$ \cite{S07}
	Furthermore, the error bound for the asymptotic limit is nearly optimal.

%\textbf{Experimental results.} We show practical implications of our theoretical insights on text and image classification tasks.

	%validate on text and image classification tasks that comparing single-task accuracies can help determine whether multi-task learning performs better than single-task learning.

>>>>

%Multi-task learning is an inductive learning mechanism to improve generalization performance using related task data.
%Many state-of-the-art results in computer vision and natural language processing are obtained using multi-task learning.
Multi-task learning has become a powerful paradigm to solve complex prediction tasks in computer vision \cite{chexnet17,ZSSGM18}, natural language processing \cite{GLUE,superglue} and numerous other areas \cite{ZY17}.
%In multi-task learning, having related task data is fundamental to its performance.
%Multi-task learning is particularly powerful when there is limited labeled data for a task to be solved, meanwhile more labeled data from different but related tasks is available.
By combining multiple information sources, it is possible to share new information among different sources in the same model \cite{C97}.
%For example, many applications in , and many other areas have been achieved by learning from multiple tasks together.
Intuitively, the performance of multi-task learning depends on the relationship of the information sources.
When the information sources are heterogeneous, negative transfer -- where multi-task learning performs worse than single-task learning -- has often been observed \cite{AP16,BS17}.
While numerous studies have sought to alleviate negative transfer when it occurs \cite{YKGLHF20}, a rigorous understanding to the contributing causes of negative transfer has remained elusive in the literature \cite{R17}.
%This phenomenon, known as \textit{negative transfer}, is fundamental to the understanding of multi-task learning.
In this work, we develop technical tools to better understand when and why negative transfer occurs for learning multiple linear regression tasks in high dimensions. % for learning from multiple linear regression tasks.
We use the tools to show theoretical and practical implications for detecting and improving negative transfer.
%We consider a setting where the target task has limited labeled data and show
%On the other hand, unless the structures across task data are well-understood, applying multi-task learning on several different datasets often result in suboptimal models (or negative transfer in more technical terms).

Identifying negative transfer requires developing tight generalization bounds for both multi-task learning and single-task learning.
In classical Rademacher or VC based theory of multi-task learning \cite{B00,AZ05,M06}, the generalization bounds are usually presented in a way so that the error goes down as more labeled data is added.
On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
More recent work has shown the benefit of learning multi-task representations for certain half-spaces \cite{MPR16} and multiple sparse regressions \cite{LPTV09,LPVT11}.
In order to rigorously understand negative transfer, the technical challenge is to develop generalization bounds that scale tightly with the qualities of the data.

@@

In this work, we consider a setting where multiple labeled linear regression tasks are available and focus on predicting a particular task whose amount of labeled data is limited.
Following Hastie et al. \cite{HMRT19} and Bartlett et al. \cite{BLLT20}, we assume that for every task $1\le i\le t$, its features are random vectors $x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
The label of $x$ is given by $y_i = x^{\top}\beta_i + \varepsilon$, where $\beta_i\in\real^p$ denotes the ground truth parameters for task $i$ and $\varepsilon$ denotes random noise with mean zero and variance $\sigma^2$.
Importantly, we assume that the training data size of the target task $t$ is a small constant $\rho_t > 1$ times $p$ to capture the need for more labeled data.
%We shall assume that each task data follows a linear model, i.e. $y_i = X_i \beta_i + \varepsilon_i$, $1\le i\le k$.
%Here $\beta_i\in\real^p$ is the model parameter for the $i$-th task.
%Each row of $X_i\in\real^{n_i\times p}$ is assumed to be drawn i.i.d. from a fixed
%distribution with covariance matrix $\Sigma_i$.

We combine all the available training data using a hard parameter sharing architecture that contains a shared body $B\in\real^{p\times r}$ for all tasks and a separate prediction head $\set{W_i \in \real^{r}}_{i=1}^t$ for each task \cite{R17,MTDNN19,WZR20}.
%    \paragraph{Different covariates:}
This corresponds to minimizing the following objective.
\begin{align}
	\label{eq_mtl}
	f(B; W_1, \dots, W_t) = \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2.
\end{align}
For a target task $t$,
let $\hat{\beta}_t^{\MTL} = B W_t$ denote the optimal multi-task estimator by solving equation \eqref{eq_mtl}.
Let $\hat{\beta}_t^{\STL}$ denote the standard linear regression estimator using task $t$ alone.
We say there is negative transfer if the test error of $\hat{\beta}_t^{\MTL}$  is smaller than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.

\textbf{Main results.}
We revisit the bias-variance tradeoff of the multi-task estimator.
Interestingly, the variance of $\hat{\beta}_t^{\MTL}$ is always smaller than that of $\hat{\beta}_t^{\STL}$, hence resulting in a positive effect of variance reduction.
The bias of $\hat{\beta}_t^{\MTL}$, which we term as \textit{model shift bias}, results in a negative effect caused by the difference between $\beta_t$ and the rest $\set{\beta_i}_{i=1}^{t-1}$.
Hence, the tradeoff between the variance reduction effect and model shift bias determines whether we observe positive or negative transfer.

To quantify the tradeoff more precisely, we develop a new technical result on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, for two random matrices $X_1$ and $X_2$ when their dimensions $p$ are large and data sizes are fixed constants times $p$.
Using the result, we provide a sharp bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for the setting of two tasks with general covariance matrices.
When there are more than two tasks, we show a similar result if they all have the same covariates.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
Finally, we apply our technical tools to the related setting of transfer learning.
We study the transfer function used by Taskonomy \cite{ZSSGM18}, which pools learnt representations from source tasks into a shared body similar to $B$ in equation \eqref{eq_mtl}.
We find that the model shift bias can be captured nicely by the orthogonal projection of $\beta_t$ to the subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
These results are presented more precisely in Section \ref{sec_main}.

The tradeoff that we developed in Section \ref{sec_main} depends on the qualities of each task dataset such as its data size and covariance matrix.
In Section \ref{sec_insight}, we use the technical results to provide guidance on when positive transfer is likely to occur and identify causes of negative transfer.
\begin{itemize}
		\item First, we provide a more precise interpretation to a folklore understanding in multi-task learning, which posits that negative transfer is caused by having dissimilar tasks.
		For an example of two tasks, we show a sharp transition from positive to negative transfer determined by the ratio of $\norm{\beta_1 - \beta_2}^2$ and a certain function of data sizes (cf. Proposition \ref{prop_dist_transition}).
		We further show that positive transfer is more likely to occur when transferring from a less noisy source task to a more noisy target task.
		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
		\item Second, we find that depending on how large $\norm{\beta_1 - \beta_2}^2$ is, adding more labeled data from the source task does not always reduce the test error of $\hat{\beta}_t^{\MTL}$ (cf. Proposition \ref{prop_data_size}).
	We connect the observation to explain a key finding of Taskonomy \cite{ZSSGM18}.
	We define the \textit{data efficiency ratio} as the smallest $\alpha\in(0, 1)$ such that if we only use an $\alpha$ fraction of labeled data, then the test error of $\hat{\beta}_t^{\MTL}$ matches that of the $\hat{\beta}_t^{\STL}$ on the entire set.
	In Proposition \ref{prop_data_efficiency}, we show that the data efficiency ratio of an illustrative example is at most $\alert{\frac {1}{2\rho_t}}$.
	In Section \ref{sec_validate}, we validate that performing multi-task learning can reduce the need for labeled data on 6 sentiment analysis tasks. %, we observe that just by using \alert{40\%} of the labeled data, the overall accuracy of multi-task learning matches that of learning every task in isolation.
		\item Finally, we find that covariate shift, i.e. having different covariance matrices, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
		We show that as $n_1 / n_2$ becomes large, the best performing source task has have the same covariance matrix as the target task (cf. Proposition \ref{prop_covariate}).
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.
\end{itemize}

%	A crucial technical tool that we develop for showing Theorem \ref{thm_main_informal} is the asymptotic limit of the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$ (for the setting of two tasks), which may be of independent interest.
	The technical tool we developed extends a well-known result in the random matrix theory literature on the trace of $(X^{\top}X)^{-1}$ \cite{S07} for a single random matrix to two random matrices.
	Our error bound for the asymptotic limit is nearly optimal, which may be of independent interest.

We show practical implications of our theory on text and image classification tasks.
	First, we provide a metric to determine positive versus negative transfer by comparing the test accuracies of single-task models.
	Second, \todo{}
	Finally, we show that as the data size between the source and target task becomes more imbalanced, aligning the covariances of the tasks becomes more beneficial.
% as the data size becomes more imbalanced between the source and target task, having mis-aligned task covariance matrices can result in suboptimal performance.
	%validate on text and image classification tasks that comparing single-task accuracies can help determine whether multi-task learning performs better than single-task learning.
